---
title: "The Democracy Gap"
subtitle: "Analyzing Gender Representation & Regime Changes in Democratic Countries"
author: "Julia Aptekar, Meghana Kolar, Sofia Jachno, & Bryce Corbett"
embed-resources: true
code-fold: true
format: html
theme: lux
---
```{r}
#| warning: false
#install.packages("tidyverse")
#install.packages("wesanderson")
#install.packages("gt")
#install.packages("broom")
options(pillar.width = Inf)
library(tidyverse)
library(wesanderson)
library(gt)
library(broom)
democracy_data <- read.csv("Combined_Democracy_Data.csv")
pal <- wes_palette("Moonrise3", n = 5)
```

Democracy is usually associated with participation, elections, and political rights, but that does not mean political power is shared equally. A country can hold elections, expand voting rights, and allow leadership changes, yet women can still be  underrepresented in government. Women remain underrepresented in many national parliaments and executive offices, even in countries that meet basic democratic standards.

Democracy itself can also shift over time. Some countries remain democratic or non-democratic for decades, while others move between the two after coups, elections, leadership changes, or broader political crises. Because of that, the global rise of democracy is not just a story of more countries becoming democratic. It is also a story of reversals, repeated transitions, and uneven political change.

## What We Want To Know

We analyzed democracy from two angles. The first is whether democratic context is associated with gender gaps in parliamentary representation and executive leadership. The second is how transitions into and out of democracy have changed the global proportion of democratic regimes over time.

## Exploring the data

To answer these questions, we used two datasets on regime type, political leadership, and women’s representation in national parliaments. These sources made it possible to compare democratic and non-democratic countries while also tracking how democracy status changes over time.

The first dataset was compiled by Christian Bjørnskov and Martin Rode, two economists who study political institutions and how governments change over time. Their goal was to create a resource for researchers studying political transitions. The dataset contains information on 192 sovereign countries and 16 self-governing territories collected annually from 1950 to 2020 on each country's regime type, democratic status, elections, suffrage, transfers of power, and more. 

To build this dataset, the authors expanded on an existing collection of political data by recording key features of each country's government, such as whether elections took place, whether those elections were considered free and fair, and whether leadership changed hands peacefully. Records of coups were verified across newspaper archives, encyclopedias, and other databases to ensure accuracy. 

Our second dataset comes from the World Bank, which tracks global progress on the UN's 17 Sustainable Development Goals, a set of goals aimed at making the world more equitable and sustainable across areas like poverty, education, and gender equality. The 5th goal focuses specifically on gender equality, and one of its key indicators is the percentage of parliamentary seats held by women in each country. This figure is compiled by the World Bank using data from the Inter-Parliamentary Union, an organization that collects official seat counts directly from legislative bodies around the world. It is available annually for most countries from 1997 to 2020, with coverage becoming more consistent in the later years of that range.

**Read more about the data here:**

[Democracy Data](https://www.christianbjoernskov.com/bjoernskovrodedata/)

[World Bank Data](https://databank.worldbank.org/source/sustainable-development-goals-(sdgs)#)

Together, these datasets let us connect democracy status to both representation within government and movement between political systems over time.

## Exploring Gender Gaps in Leadership and Representation

Democracy status gives one way to compare gender representation across countries. Since democracies are associated with voting rights, elections, and political participation, the question is whether those features also show up in who holds political power. Women’s representation in national parliaments gives one view of that gap, while presidents and monarchs show gender differences at the highest levels of leadership.

These measures capture different parts of political power. Parliament reflects representation within national lawmaking bodies, while executive leadership shows who reaches the top positions in government. Looking at both helps show whether democracy status is connected to broader representation, executive leadership, or both.

```{r}
#| warning: false
democracy_data <- democracy_data |>
  mutate(president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950s",
       ifelse(year < 1970, "1960s",
       ifelse(year < 1980, "1970s",
       ifelse(year < 1990, "1980s",
       ifelse(year < 2000, "1990s",
       ifelse(year < 2010, "2000s",
                           "2010s")))))))

summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |> 
    filter(!is.na(democracy_data[[group_var]])) |>
    # grouping data by specified grouping variable
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
              # calculating proportions of female and male presidents
              prop_female_president = mean(
                ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
              prop_male_president = mean(
                ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
              # calculating difference in proportions of male presidents and female presidents (male - female)
              gap_president_prop = prop_male_president - prop_female_president,
              # calculating the average accession ages for female and male presidents
              avg_accession_age_female_president = mean(
                president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
              avg_accession_age_male_president = mean(
                president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
              # calculating difference in average accession ages of male and female presidents (male - female)
              gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
              
              # calculating proportions of female and male monarchs
              prop_female_monarch = mean(
                ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
              prop_male_monarch = mean(
                ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
              # calculating difference in proportions of male monarchs and female monarchs (male - female)
              gap_monarch_prop = prop_male_monarch - prop_female_monarch,
              # calculating the average accession ages for female and male monarchs
              avg_accession_age_female_monarch = mean(
                monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
              avg_accession_age_male_monarch = mean(
                monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
              # calculating difference in average accession ages of male and female monarchs (male - female)
              gap_monarch_accession_age =
                avg_accession_age_male_monarch - avg_accession_age_female_monarch,
              .groups = "drop")
}

plotMe <- summarise_leadership_gender_gap("year")
```

We began by comparing the percentages of parliamentary seats held by women in democratic and non-democratic countries. 

```{r}
#| fig-alt: This is a boxplot depicting the proportion of women in parliament, while faceting on the democracy status of the country. The mean proportion of women holding seats in non-democratic countries is 0.16 and 0.177 for democratic countries. The median proportion of women holding seats in non-democratic countries is 0.14 and 0.158 for democratic countries.

pal0 <- wes_palette("Moonrise3", n = 2)  
democracy_data %>%
  filter(!is.na(is_democracy)) %>%
  filter(!is.na(prop_in_par)) %>%
  mutate(is_democracy = factor(is_democracy, labels = c("Non-Democracy", "Democracy"))) %>%
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  theme(text = element_text(size = 11, family = "serif")) +
  geom_boxplot(fill = pal0,              
    color = "black",          
    outlier.color = "black") +
  stat_summary(fun = mean,
    geom = "crossbar",
    color = "black",
    width = 0.75,
    linetype = "dotted",
    linewidth = .2) +
  stat_summary(fun = mean,
    geom = "text",
    aes(label = paste0("Mean: ", round(after_stat(y), 1),"%")),
    vjust = -1,
    color = "black") +
  stat_summary(fun = median,
    geom = "text",
    aes(label = paste0("Median: ", round(after_stat(y), 1),"%")),
    vjust = 1.5,
    color = "black") +
  labs(title = "Women's Representation in National Parliament by Democracy Status",
    x = "Democracy Status",
    y = "Percentage of National Parliament Seats Held by Women") + 
  theme(plot.title = element_text(size = 5),
    axis.title = element_text(size = 5)) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_minimal()
```

On average, democracies tend to have slightly higher percentages of women in parliament, with a median of 15.8% compared to 14.0% in non-democratic countries, and a mean of 17.7% versus 16.0%. That said, the two groups look fairly similar overall, with significant overlap between them and wide variation within each group. Non-democratic countries do show a few unusually high values, including two countries where women held more than 60% of parliamentary seats.

Because the boxplot showed a small difference between democracies and non-democracies, we used a Welch two-sample t-test to check whether the difference in average women’s parliamentary representation was statistically significant.

```{r}
result <- t.test(prop_in_par ~ is_democracy, data = democracy_data)
table_result <- tidy(result)

table_result |>
  select(estimate1, estimate2, statistic, p.value, conf.low, conf.high, parameter) |>
  gt() |>
  tab_header(title = md("**Proportion in Parliament by Democracy Status**"), subtitle = "Welch Two-Sample t-Test") |>
  cols_label(estimate1 = md("Mean<br>(Non-Democracy)"),
    estimate2 = md("Mean<br>(Democracy)"),
    statistic = md("*t*-statistic"),
    p.value = md("*p*-value"),
    conf.low = "Lower Bound",
    conf.high = "Upper Bound",
    parameter = md("*df*")) |>
  fmt_number(columns  = c(estimate1, estimate2, statistic, conf.low, conf.high),decimals = 3) |>
  fmt_number(columns  = parameter,
    decimals = 1) |>
  fmt_scientific(columns = p.value,
    decimals = 3) |>
  cols_align(align = "center",
    columns = everything()) |>
  tab_style(style = cell_fill(color = "#FAD77B"),
    locations = cells_body(columns = p.value)) |>
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_body(columns = p.value)) |>
  tab_spanner(label = "Group Means",
    columns = c(estimate1, estimate2)) |>
  tab_spanner(label = "Confidence Interval",
    columns = c(conf.low, conf.high)) |>
  tab_footnote(footnote = "Highlighted cell indicates a statistically significant result.", locations = cells_column_labels(columns = p.value)) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

The t-statistic is -4.591 and the p-value is very small. Also, the confidence interval does not include zero, so we reject the null hypothesis and conclude that there is a statistically significant difference in the mean proportion of seats held by women between democratic and non-democratic countries. The result suggests that democracies and non-democracies differ in average women’s parliamentary representation, although the size of the difference is modest.

To better see how democracy status relates to gender gaps in executive leadership, we compared the proportions of male and female presidents and monarchs in democratic and non-democratic countries.

```{r}
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n, prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop) |>
  gt() |>
  tab_header(title = md("**Gender Gaps in Executive Leadership by Democracy Status**"),
    subtitle = "Average Percentages of Female and Male Leaders") |>
  tab_spanner(label = "Presidents",
    columns = c(prop_female_president, prop_male_president, gap_president_prop)) |>
  tab_spanner(label = "Monarchs",
    columns = c(prop_female_monarch, prop_male_monarch, gap_monarch_prop)) |>
  cols_label(is_democracy = "Democracy Status",
    n = "Sample Size", prop_female_president = "Female",
    prop_male_president = "Male",
    gap_president_prop = "Gap",
    prop_female_monarch = "Female",
    prop_male_monarch = "Male",
    gap_monarch_prop = "Gap") |>
  fmt_percent(columns = c(prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  cols_align(align = "center",
    columns = everything()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_president_prop,
      rows = gap_president_prop > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_president_prop,
      rows = gap_president_prop < 0)) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_monarch_prop,
      rows = gap_monarch_prop > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_monarch_prop,
      rows = gap_monarch_prop < 0)) |>
  tab_footnote(footnote  = "Positive values (shaded blue) indicate males are higher; negative values (shaded pink) indicate females are higher. Gap is calculated as M - F.",
    locations = cells_column_labels(columns = c(gap_president_prop, gap_monarch_prop))) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

Presidents are overwhelmingly male in both democratic and non-democratic countries. Female presidents account for just 0.5% of presidents in non-democracies, compared to 5.6% in democracies. That is a noticeable difference, but it also shows how rare female presidents are overall. Even in democracies, more than 94% of presidents are men.

Monarchies tell a different story. In non-democratic countries, monarchies are still heavily male, with men outnumbering women by about 84 percentage points. In democratic countries, the split is much closer, and women actually make up a slightly larger share of monarchs than men. So while democracy status seems to be linked with smaller gender gaps in executive leadership, the pattern depends on the type of leader. 

We then looked at accession ages to see whether gender differences also appear in when leaders first take power.

```{r}
# new with center align
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
         avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age,
         avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age) |>
  gt() |>
  tab_header(title = md("**Gender Gaps in Accession Age by Democracy Status**"),
    subtitle = "Average Accession Ages for Female and Male Leaders") |>
  tab_spanner(label = "Presidents",
    columns = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age)) |>
  tab_spanner(label = "Monarchs",
    columns = c(avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age)) |>
  cols_label(is_democracy = "Democracy Status",
    n = "Sample Size",
    avg_accession_age_female_president = "Female",
    avg_accession_age_male_president = "Male",
    gap_president_accession_age = "Gap",
    avg_accession_age_female_monarch  = "Female",
    avg_accession_age_male_monarch = "Male",
    gap_monarch_accession_age = "Gap") |>
  fmt_number(columns  = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age, avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  cols_align(align = "center",
    columns = everything()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_president_accession_age,
      rows = gap_president_accession_age > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_president_accession_age,
      rows  = gap_president_accession_age < 0)) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_monarch_accession_age,
      rows = gap_monarch_accession_age > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_monarch_accession_age,
      rows = gap_monarch_accession_age < 0)) |>
  tab_footnote(footnote  = "Positive values (shaded blue) indicate males are higher; negative values (shaded pink) indicate females are higher. Gap is calculated as M - F.",
    locations = cells_column_labels(columns = c(gap_president_accession_age, gap_monarch_accession_age))) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

Among presidents, the biggest age gap appears in non-democratic countries. Female presidents in non-democracies took office at an average age of 62.4, compared with 48.6 for male presidents. In democracies, the gap is much smaller, with female presidents taking office at 54.2 on average and male presidents at 57.0.

For monarchs, the pattern is more consistent across democracy status. Male monarchs came to power later than female monarchs in both groups, by about 7 years in non-democratic countries and almost 8 years in democracies. This makes monarchs different from presidents: the presidential age gap changes a lot depending on democracy status, while the monarch age gap stays fairly similar.

The table also shows a clear age difference between presidents and monarchs overall. Presidents tend to come to power much later in life, mostly in their 50s and 60s, while monarchs usually take power in their late 20s or 30s.

To see how these age gaps have shifted over time, we constructed a ribbon plot showing the difference between average accession ages for male and female presidents and monarchs by decade. Because there were no recorded female presidents before 1980, we limited our visual to years from 1980 onward.

```{r}
#| warning: false
#| fig-alt: This is a ribbon plot, depicting the differences in average accession ages of presidents and monarchs by gender. There are four lines on the graph, one for each gender and leader type combination. The x-axis is years 1980-2020, and the y-axis shows ages 20-70. 

plotMe %>%
  filter(year >= 1980) %>%
  ggplot(aes(x = year)) +
  geom_ribbon(data = plotMe, aes(x = year,
                                 ymin = avg_accession_age_female_president,
                                 ymax = avg_accession_age_male_president),
              fill = "#D3D3D3", alpha = 0.5) +
  geom_ribbon(data = plotMe, mapping = aes(x = year,
                                           ymin = avg_accession_age_female_monarch,
                                           ymax = avg_accession_age_male_monarch),
              fill = "#D3D3D3", alpha = 0.5) +
  geom_line(aes(y = avg_accession_age_female_president, color = "Female President"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_president, color = "Male President"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_female_monarch, color = "Female Monarch"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_monarch, color = "Male Monarch"), linewidth = 1.5) +
  scale_color_manual(name = "Leader Type",
    values = c("Female President" = pal[3],
      "Male President"= pal[2],
      "Male Monarch" = pal[4],
      "Female Monarch" = pal[1]),
    breaks = c("Female President", "Male President", "Male Monarch", "Female Monarch")) +
  scale_x_continuous(limits = c(1980, 2020)) +
  scale_y_continuous(limits = c(20, 70)) +
  labs(title = "Average Accession Ages of Presidents & Monarchs by Gender", x = "Year", y = "Average Accession Age") +
  theme_minimal()
```

Among presidents, the accession age gap is small and relatively stable. Female presidents were older on average than male presidents until around 1995, when the average accession age for male presidents overtook that of female presidents. The opposite change occurred around 2005, and afterwards female presidents remain older on average than male presidents.

In contrast, monarchs show a much larger and growing gap. Male monarchs tend to take the throne at older ages, while female monarchs are generally younger, with their ages even decreasing in more recent years. Overall, gender disparities in age at taking office are minimal in presidential systems but much more pronounced in monarchical ones.

After looking at democracy status, we also wanted to see how gender gaps in executive leadership changed over time. Grouping the data by decade shows whether women became more common among presidents and monarchs from 1950 to 2020.

```{r}
#| results: asis
group_vars <- c("decade")
group_labels <- c("Decade")

walk2(group_vars, group_labels, function(var, label) {
  tbl <- summarise_leadership_gender_gap(var) |>
    select(all_of(var), n, prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop)
  
  print(tbl |> gt() |>
    tab_header(title = md(paste0("**Gender Gaps by ", label, "**")),
               subtitle = "Percentages of Female and Male Leaders") |>
    tab_spanner(label = "Presidents",
      columns = c(prop_female_president, prop_male_president, gap_president_prop)) |>
    tab_spanner(label = "Monarchs",
      columns = c(prop_female_monarch, prop_male_monarch, gap_monarch_prop)) |>
    cols_label(!!sym(var) := label,
      n = "Sample Size",
      prop_female_president = "Female",
      prop_male_president = "Male",
      gap_president_prop = "Gap (M - F)",
      prop_female_monarch = "Female",
      prop_male_monarch = "Male",
      gap_monarch_prop = "Gap (M - F)") |>
    fmt_percent(columns = c(prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop),
      decimals = 1) |>
    fmt_integer(columns = n) |>
    cols_align(align = "center",
      columns = everything()) |>
    tab_options(table.width = pct(100),
      table.font.size = px(18),
      heading.title.font.size = px(26),
      heading.subtitle.font.size = px(18),
      column_labels.font.size = px(18),
      row_group.font.size = px(18),
      data_row.padding = px(10),
      column_labels.padding = px(8),
      heading.padding = px(8)) |>
    opt_table_font(font = "serif") |>
    tab_style(style = cell_fill(color = "#D3D3D3"),
      locations = cells_column_spanners()) |>
    tab_style(style = cell_fill(color = "#D3D3D3"),
      locations = cells_column_labels()) |>
    tab_style(style = cell_fill(color = "#D3D3D3"),
      locations = cells_title()))
})
```

Until the 1970s, there were no female presidents. The percentage of female presidents increased quickly over the decades, ending at 7% in the 2010s. The percentage of female monarchs also increased over the decades, but actually began decreasing in the 2000s and 2010s. 

Across the gender representation measures, democracy status matters, but only to a point. The t-test shows that the difference in average representation is statistically significant, meaning it is unlikely to be due to random variation alone. Still, the boxplot shows a lot of overlap between the two groups, so democracy status does not fully lead to higher women’s representation.

Executive leadership shows an even larger gender gap. Female presidents are more common in democracies than in non-democracies, but presidents are overwhelmingly male in both groups. Monarchs show a different pattern: non-democratic monarchies are heavily male, while democratic monarchies are much closer to balanced. Overall, democracy status seems to affect the number of female leaders, but it does not remove gender inequality in political leadership.

Gender representation shows one way of how democracy relates to political power. Another way is whether democratic rule itself lasts, since countries can move into and out of democracy over time.

## Exploring The Global Effect of Democratic Transitions

Democracy status can change over time. Some countries remain democratic or non-democratic for decades, while others switch between the two after elections, coups, leadership changes, or conflicts. Because of this, the global rise in democracy is not just about how many countries are democratic in a given year, but also about how often countries move into and out of democratic rule.

Tracking these transitions from 1950 to 2020 gives a clearer view of where the overall trend comes from. It shows which periods had the most movement toward democracy, when reversals happened, and which countries changed status most often.

We first examined the overall trend of the percentage of democratic countries. 

```{r}
#arranging by country and year
democracy_data <- democracy_data |> 
  arrange(country, year)

#creating new columns to the data, and assigning them to 0 to start with. this is an indicator variable, with 0 being no transition, and 1 meaning there is a transition. for example, if for a country 2000 is a 0 for the to_democracy variable and 2001 there is a 1, there was a transition to democracy in 2001. 
democracy_data$to_democracy <- 0
democracy_data$from_democracy <- 0

#to iterate through rows to find out which ones have transitions. starting at 2 because we are comparing to the previous row (2-1 = 1)
for (i in 2:nrow(democracy_data)) {
  #checking if the two countries that are next to each other in the data are the same 
  if (democracy_data$country[i] == democracy_data$country[i-1]) { 
    #assigning them to current and previous labels
    current <- democracy_data$is_democracy[i]
    previous <- democracy_data$is_democracy[i - 1]
    #making sure there are no na values for the is_democracy variable
    if (!is.na(current) && !is.na(previous)) {
      #if this year there is democracy and the previous year there wasn't, it is marked as a transition to democracy
      if (current && !previous) {
        democracy_data$to_democracy[i] <- 1
      }
      #opposite for this one 
    if (!current && previous) {
        democracy_data$from_democracy[i] <- 1
      }
    }
  }
}

#first grouping by year, and then summarizing the proportion of democracy, and transitions in and out of democracy
summary_data <- democracy_data |>
  group_by(year) |>
  summarize(
    prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE))

summary_ten_yrs <- democracy_data |>
  group_by(decade) |>
  summarize(
    prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE),
    .groups = "drop")

ggplot(summary_data, aes(x = year, y = prop_democracy)) +
  geom_point(color = pal[2], size = 2) +
  geom_line(color=pal[2], linewidth=0.75) +
  scale_y_continuous(breaks=seq(0,1,0.2), limits=c(0,1),
                     labels = scales::percent_format()) +
  labs(title = "Percentage of Democratic Countries Over Time",
    x = "Year",
    y = "Percentage of Democratic Countries") +
  scale_x_continuous(breaks=seq(1950,2020,10)) +
  theme_minimal()
```

In 1950, the percentage of democratic countries was at about 25%, before rising gradually through the 1960s. The percentage then stayed fairly stable around the mid-30% range through the 1970s and early 1980s. The largest increase happened around the late 1980s and early 1990s, when the percentage increased sharply to nearly 60%. After that, democracy continued to grow more slowly, reaching the mid-60% range in the 2010s before leveling off slightly by 2020.

To see when the biggest shifts happened, we looked at how many countries entered or left democracy in each period.

```{r}
#| results: asis
summary_ten_yrs |>
  select(decade, transitions_in, transitions_out) |>
  gt() |>
  tab_header(title = md("**Democracy Transitions by Year**"),
    subtitle = "Transitions In and Out of Democracy") |>
  cols_label(decade = "Decade",
    transitions_in = "In",
    transitions_out =  "Out") |>
  fmt_integer(columns = c(transitions_in, transitions_out)) |>
  cols_align(align = "center",
    columns = everything()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

The 1990s stand out as the most transformative decade, with 39 transitions into democracy and only 11 out, reflecting the wave of democratization that followed the end of the Cold War. In contrast, the 1970s were the most stagnant period, with transitions in and out nearly balanced at 16 and 15 respectively and virtually no net change in the global proportion. 

Our area plot below displays the counts of democratic transitions over the years visually.

```{r}
#| fig-alt: The area plot below shows the count of democractic transitions in and out, where the pink represents a country gaining a democratic status, and the blue represents a country losing a democratic status throughout the years.

summary_data_long <- summary_data |>
  pivot_longer(cols = c(transitions_in, transitions_out),
    names_to = "type",
    values_to = "count")

ggplot(summary_data_long, aes(x = year, y = count, fill = type)) +
  geom_area(alpha = 0.75) +
  scale_fill_manual(values = c(transitions_in = pal[2],
      transitions_out = pal[1]),
    name = "Transition Direction",
    labels = c(transitions_in = "Transitions In",
      transitions_out = "Transitions Out")) +
  labs(title = "Global Democratic Transitions Over Time",
    x = "Year",
    y = "Count of Transitions") +
  scale_y_continuous(limits=c(0,12), breaks=seq(0,12,2)) +
  scale_x_continuous(limits=c(1950,2020), breaks=seq(1950,2020,10)) +
  theme_minimal()
```

The area plot shows that democratic transitions happened in bursts rather than at a steady pace. Transitions into democracy were often more common than transitions out, especially during the 1990s, but reversals continued throughout the full time period. This supports the idea that democratization was not a smooth or one-directional process.

We also wanted to identify which countries had the most democratic transitions.

```{r}
topn_countries <- function(start_year = 1950, end_year = 2020, num_countries = 10) {
  if (start_year < 1950 | end_year > 2020 | start_year > end_year) {
    stop("Year range must be between 1950 and 2020")
  }
  
  data <- democracy_data |>
    filter(year >= start_year & year <= end_year) |>
    arrange(country_name, year) |>
    group_by(country_name) |>
    mutate(prev_year = lag(is_democracy),
      to_democracy = ifelse(is_democracy == 1 & prev_year == 0, 1, 0),
      from_democracy = ifelse(is_democracy == 0 & prev_year == 1, 1, 0)) |>
    ungroup()
  
  country_table <- data |>
    group_by(country_name) |>
    summarize(transitions_in = sum(to_democracy, na.rm = TRUE),
      transitions_out = sum(from_democracy, na.rm = TRUE),
      total_transitions = transitions_in + transitions_out,
      .groups = "drop") |>
    arrange(desc(total_transitions)) |>
    slice_head(n = num_countries)
  
  return(country_table)
}
country_transitions <- topn_countries(1950, 2020, 10)
country_transitions |>
  gt() |>
  tab_header(title = md("**Top 10 Countries by Democratic Transitions**"),
    subtitle = "Total Transitions Into and Out of Democracy") |>
  cols_label(country_name = "Country",
    transitions_in = "Transitions In",
    transitions_out = "Transitions Out",
    total_transitions = "Total Transitions") |>
  fmt_integer(columns = c(transitions_in, transitions_out, total_transitions)) |>
  cols_align(align = "center",
    columns = everything()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

Argentina and Thailand experienced the highest number of democratic transitions between 1950 and 2020, followed by Ghana, Pakistan, and Peru. Having more repeated shifts between democratic and non-democratic rule indicates political instability.

Lastly, our table earlier showed that the most transitions into democracy happened between 1980-2010. We created another table focusing in on the top ten countries in this time frame.

```{r}
country_transitions2 <- topn_countries(1980, 2010, 10)
country_transitions2 |>
  gt() |>
  tab_header(title = md("**Top 10 Countries by Democratic Transitions**"),
    subtitle = "Total Transitions Into and Out of Democracy from 1980-2010") |>
  cols_label(country_name = "Country",
    transitions_in = "In",
    transitions_out = "Out",
    total_transitions = "Total Transitions") |>
  fmt_integer(columns = c(transitions_in, transitions_out, total_transitions)) |>
  cols_align(align = "center",
    columns = everything()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

In this table and the previous one, we see that the countries listed all have four or fewer transitions in or out of democracy in the respective time frames. Although this is not a very large number, even a small number of transitions indicates meaningful political instability, since transitions between democratic and non-democratic systems often reflect major shifts in leadership and ideology. This suggests that a key part of democratization is not only becoming democratic, but also maintaining democratic rule over time. 

## Where the Gap Remains

Overall, we find that democracy is related to gender representation and political stability, but not always in simple ways. Democratic countries had slightly higher average proportions of women in parliament than non-democratic countries, and our t-test showed that this difference was statistically significant. However, the actual difference in means was fairly small, suggesting that democracy alone does not guarantee equal representation. Female presidents became more common over time, but male presidents still made up the large majority of presidential leaders across every decade. Accession age gaps were smaller and more stable among presidents, while monarchs showed larger differences by gender.

We also saw that global democratization has been unpredictable. The percentage of democratic countries increased substantially from 1950 to 2020, especially around the late 1980s and early 1990s, but this growth was not steady. Countries continued to move both into and out of democracy over time, showing that democratization includes both progress and reversal. The country-level transition tables suggest that even a small number of regime changes can reflect meaningful political instability.

Becoming democratic is associated with some gains in representation, but the data do not show democracy as a simple solution to gender inequality or political instability. Women remain underrepresented in many leadership roles, and countries continue to move into and out of democratic rule. The democracy gap, then, is not only about whether a country is democratic, but also about who gains power within that system and whether the system lasts.

