---
title: "Democracy & Gender in Leadership"
format:
  dashboard:
    orientation: columns
    theme: flatly
    css: dashboard-styles.css
    logo: logo.png
    logo-alt: "Democracy and Gender in Leadership project logo."
embed-resources: true
nav-buttons:
  - icon: github
    href: https://github.com/BryceCorbett/Geopolitical-Analysis
    aria-label: GitHub
  - icon: linkedin
    href: https://www.linkedin.com
    aria-label: LinkedIn
  - icon: instagram
    href: https://www.instagram.com
    aria-label: Instagram
  - icon: envelope
    href: mailto:yourteam@email.com
    aria-label: Email
---
```{r}
#| echo: false
#| message: false
#| warning: false

options(pillar.width = Inf)
library(tidyverse)
library(wesanderson)
library(gt)
library(broom)

democracy_data <- read.csv("Combined_Democracy_Data.csv")

democracy_data <- democracy_data |>
  mutate(
    president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession   = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950-1959",
             ifelse(year < 1970, "1960-1969",
             ifelse(year < 1980, "1970-1979",
             ifelse(year < 1990, "1980-1989",
             ifelse(year < 2000, "1990-1999",
             ifelse(year < 2010, "2000-2009",
                                 "2010-2020"))))))
  )

summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |>
    filter(!is.na(democracy_data[[group_var]])) |>
    group_by(across(all_of(group_var))) |>
    summarise(
      n = n(),
      prop_female_president = mean(ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
      prop_male_president   = mean(ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
      gap_president_prop    = prop_male_president - prop_female_president,
      avg_accession_age_female_president = mean(president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
      avg_accession_age_male_president   = mean(president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
      gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
      prop_female_monarch = mean(ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
      prop_male_monarch   = mean(ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
      gap_monarch_prop    = prop_male_monarch - prop_female_monarch,
      avg_accession_age_female_monarch = mean(monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
      avg_accession_age_male_monarch   = mean(monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
      gap_monarch_accession_age = avg_accession_age_male_monarch - avg_accession_age_female_monarch,
      .groups = "drop"
    )
}

plotMe    <- summarise_leadership_gender_gap("year")
pal       <- wes_palette("Moonrise3")

blue_dark   <- "#5b8db8"
blue_mid    <- "#89afd4"
blue_light  <- "#bdd4e7"
blue_pale   <- "#ddeaf5"
pink_accent <- "#e8a0b0"
pink_pale   <- "#f5d5dc"

theme_set(theme_minimal(base_size = 13, base_family = "Georgia"))
```

## Column {width=58%}

::: {.card title="Visualizations"}
::: {.panel-tabset .tabset-sync}
## Women in Parliament

```{r}
#| fig-height: 5.2
democracy_data |>
  filter(!is.na(is_democracy), !is.na(prop_in_par)) |>
  mutate(is_democracy = factor(is_democracy, labels = c("Non-Democracy", "Democracy"))) |>
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  geom_boxplot(
    fill          = c(blue_light, blue_dark),
    color         = blue_dark,
    outlier.color = blue_mid
  ) +
  stat_summary(
    fun      = mean,
    geom     = "crossbar",
    color    = blue_dark,
    width    = 0.75,
    linetype = "dotted",
    linewidth = 0.3
  ) +
  stat_summary(
    fun   = mean,
    geom  = "text",
    aes(label = paste0("Mean: ", round(after_stat(y), 1), "%")),
    vjust = -1,
    color = "white",
    size  = 3.5
  ) +
  stat_summary(
    fun   = median,
    geom  = "text",
    aes(label = paste0("Median: ", round(after_stat(y), 1), "%")),
    vjust = 1.5,
    color = "white",
    size  = 3.5
  ) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(x = "Democracy Status", y = "% of Parliament Seats Held by Women") +
  theme(
    panel.grid.minor = element_blank(),
    axis.title       = element_text(color = blue_dark),
    axis.text        = element_text(color = blue_dark)
  )
```

## Accession Ages Over Time

No female presidents were recorded before 1980, so this plot begins there.

```{r}
#| fig-height: 5.2
#| warning: false
plotMe |>
  filter(year >= 1980) |>
  ggplot(aes(x = year)) +
  geom_ribbon(
    aes(ymin = avg_accession_age_female_president, ymax = avg_accession_age_male_president),
    fill = blue_pale, alpha = 0.8
  ) +
  geom_ribbon(
    aes(ymin = avg_accession_age_female_monarch, ymax = avg_accession_age_male_monarch),
    fill = pink_pale, alpha = 0.8
  ) +
  geom_line(aes(y = avg_accession_age_female_president, color = "Female President"), linewidth = 1.4) +
  geom_line(aes(y = avg_accession_age_male_president,   color = "Male President"),   linewidth = 1.4) +
  geom_line(aes(y = avg_accession_age_female_monarch,   color = "Female Monarch"),   linewidth = 1.4) +
  geom_line(aes(y = avg_accession_age_male_monarch,     color = "Male Monarch"),     linewidth = 1.4) +
  scale_color_manual(
    name   = "Leader Type",
    values = c(
      "Female President" = pink_accent,
      "Male President"   = blue_dark,
      "Female Monarch"   = pink_accent,
      "Male Monarch"     = blue_mid
    ),
    breaks = c("Female President", "Male President", "Female Monarch", "Male Monarch")
  ) +
  scale_x_continuous(limits = c(1980, 2020)) +
  scale_y_continuous(limits = c(20, 70)) +
  labs(x = "Year", y = "Average Accession Age", color = "Leader Type") +
  theme(
    panel.grid.minor  = element_blank(),
    legend.position   = "bottom",
    legend.direction  = "horizontal",
    legend.background = element_rect(fill = "white", color = blue_light),
    legend.text       = element_text(size = 9, color = blue_dark),
    legend.title      = element_text(size = 9, color = blue_dark),
    axis.title        = element_text(color = blue_dark),
    axis.text         = element_text(color = blue_dark)
  )
```

:::
:::

## Column {width=42%}

::: {.card title="Tables"}
::: {.panel-tabset .tabset-sync}
## Women in Parliament

```{r}
result       <- t.test(prop_in_par ~ is_democracy, data = democracy_data)
table_result <- tidy(result)

table_result |>
  select(estimate1, estimate2, statistic, p.value, conf.low, conf.high, parameter) |>
  gt() |>
  tab_header(
    title    = md("**Welch Two-Sample t-Test**"),
    subtitle = "Proportion of Seats Held by Women"
  ) |>
  cols_label(
    estimate1 = md("Mean<br>(Non-Dem.)"),
    estimate2 = md("Mean<br>(Dem.)"),
    statistic = md("*t*"),
    p.value   = md("*p*-value"),
    conf.low  = "CI Low",
    conf.high = "CI High",
    parameter = md("*df*")
  ) |>
  fmt_number(columns  = c(estimate1, estimate2, statistic, conf.low, conf.high), decimals = 3) |>
  fmt_number(columns  = parameter, decimals = 1) |>
  fmt_scientific(columns = p.value, decimals = 3) |>
  tab_style(
    style     = list(cell_fill(color = pink_pale), cell_text(weight = "bold", color = blue_dark)),
    locations = cells_body(columns = p.value)
  ) |>
  tab_style(
    style     = cell_fill(color = blue_pale),
    locations = list(cells_column_spanners(), cells_column_labels(), cells_title())
  ) |>
  tab_style(
    style     = cell_text(color = blue_dark),
    locations = list(cells_column_labels(), cells_title(), cells_body())
  ) |>
  tab_footnote(
    footnote  = "Highlighted p-value is statistically significant (p < 0.05).",
    locations = cells_column_labels(columns = p.value)
  ) |>
  tab_options(
    table.font.size                = px(13),
    data_row.padding               = px(5),
    table.background.color         = "white",
    column_labels.background.color = blue_pale
  ) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

::: {.card style="background-color: #f7fbff; border: 1px solid #bdd4e7; padding: 1rem; margin-top: 1rem;"}
On average, democracies have slightly higher women's parliamentary representation — a mean of **17.7%** versus **16.0%** in non-democracies. While the difference appears small, the t-test confirms it is statistically significant (p < 0.001). Both groups show wide variation, and a handful of non-democratic countries exceed 60% representation.
:::

## Accession Ages Over Time

```{r}
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
         avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age,
         avg_accession_age_female_monarch,   avg_accession_age_male_monarch,   gap_monarch_accession_age) |>
  gt() |>
  tab_header(
    title    = md("**Gender Gaps in Accession Age**"),
    subtitle = "Average Age When Leaders First Take Office"
  ) |>
  tab_spanner(label = "Presidents",
    columns = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age)) |>
  tab_spanner(label = "Monarchs",
    columns = c(avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age)) |>
  cols_label(
    is_democracy                       = "Status",
    n                                  = "N",
    avg_accession_age_female_president = "Female",
    avg_accession_age_male_president   = "Male",
    gap_president_accession_age        = "Gap (M−F)",
    avg_accession_age_female_monarch   = "Female",
    avg_accession_age_male_monarch     = "Male",
    gap_monarch_accession_age          = "Gap (M−F)"
  ) |>
  fmt_number(columns = c(avg_accession_age_female_president, avg_accession_age_male_president,
                          gap_president_accession_age, avg_accession_age_female_monarch,
                          avg_accession_age_male_monarch, gap_monarch_accession_age),
             decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(
    style     = cell_fill(color = blue_pale),
    locations = list(cells_column_spanners(), cells_column_labels(), cells_title())
  ) |>
  tab_style(
    style     = cell_text(color = blue_dark),
    locations = list(cells_column_labels(), cells_title(), cells_body())
  ) |>
  tab_style(
    style     = cell_fill(color = pink_pale),
    locations = cells_body(
      columns = c(gap_president_accession_age, gap_monarch_accession_age),
      rows    = gap_president_accession_age > 0
    )
  ) |>
  tab_style(
    style     = cell_fill(color = blue_light),
    locations = cells_body(
      columns = c(gap_president_accession_age, gap_monarch_accession_age),
      rows    = gap_president_accession_age < 0
    )
  ) |>
  tab_footnote(
    footnote  = "Blue = males higher; pink = females higher.",
    locations = cells_column_labels(columns = c(gap_president_accession_age, gap_monarch_accession_age))
  ) |>
  tab_options(
    table.font.size                = px(13),
    data_row.padding               = px(5),
    table.background.color         = "white",
    column_labels.background.color = blue_pale
  ) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

::: {.card style="background-color: #f7fbff; border: 1px solid #bdd4e7; padding: 1rem; margin-top: 1rem;"}
In non-democratic countries, female presidents take office nearly **14 years older** than male presidents on average. In democracies, that gap shrinks to under 3 years. For monarchs, the pattern reverses — female monarchs tend to ascend younger, and the gap has been growing since 2000.
:::

:::
:::
