---
title: "Step 3 Project 1"
format: html
editor: visual
---

Code For Github Adding the nominal gdp variable

```{r}
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
library(readxl)
```

```{r}
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

```{r}
#| eval: True
my_key<-"jm6LTqDYOeb7igZXbOxfolhV4eNVaznj2yQqj2EC"
```

```{r}
get_gdp <- function(country, key = my_key)
  {
  url <- glue("https://api.api-ninjas.com/v1/gdp?country={country};X-Api-Key={key}")
  my_data <- safely(fromJSON)(url)
  
  parsed <- my_data$result
  
  if (is.null(parsed) || length(parsed$year) == 0) {
    return(NULL)
  } else {
    return(
      data.frame(
        country = country,
        year = my_data$result$year,
        gdp_nominal = my_data$result$gdp_nominal
      )
    )
  }
  }
```

```{r}
#Data cleaning: Renaming country_code column to just country to make it easier to join the datasets
democracy_data <- democracy_data %>% 
  rename(country = country_code) 
```

```{r}
gdp_data <- democracy_data |> 
  distinct(country) |>
  pull(country) |> 
  map(get_gdp) |> 
  bind_rows() |>
  filter(year <= 2025)
```

```{r}
democracy_data <- democracy_data |>
  left_join(gdp_data, by = c("country", "year"))
```

## New Data from World Bank

The World Bank’s SDG database is a subset of its broader World Development Indicators, organized specifically around the UN’s 17 Sustainable Development Goals. It provides country-year panel data (typically yearly) covering economic, social, environmental, and institutional statistics aligned with each goal. The SDGs are separated by topic (e.g., poverty, education, gender equality).

The proportion of women in national parliaments variable is under the 5th SDG. The 5th SDG is "Achieve gender equality and empower all women and girls." This variable measures the percentage of seats held by women in a country's national parliament, typically focusing on the lower or single legislative chamber. It is reported annually and serves as an indicator of gender equality in politics.

This proportion is available for most countries from 1997-2020. There are missing values for certain years and countries, but the proportion is more available during the later years of that range.

## Links

-   [SDG 5 description](https://sdgs.un.org/goals/goal5)
-   [Where we found the data](https://databank.worldbank.org/source/sustainable-development-goals-(sdgs)#)
-   [Variable description from World Bank](https://databank.worldbank.org/metadataglossary/world-development-indicators/series/SG.GEN.PARL.ZS)

## Data Cleaning

Below, we read in the data and do some data cleaning. We removed some columns because certain years had no data for any countries, and there was an empty column that was created from downloading the data. Lastly, we renamed the country column.

```{r}
womenPar0 <- read_excel("womenPar.xlsx", skip=2)
womenPar0 <- womenPar0 %>%
  select(-2:-8, -11) %>% # removing columns with no data and empty column
  rename(country_name = ...1) # renaming the country column
womenPar0
```

This data frame is in a wide format, so we used pivot_longer() to pivot the data into long format to match our democracy_data.

```{r}
womenPar <- womenPar0 %>%
  pivot_longer(cols = -country_name, # keep country column the same
    names_to = "year", # creating new column for years
    values_to = "prop_in_par") # creating new column for the proportion data
womenPar
```

```{r}

democracy_data |>
  filter(!is.na(gdp_nominal.x), !is.na(is_democracy)) |>
  group_by(is_democracy) |>
  summarise(
    mean_gdp = round(mean(gdp_nominal.x, na.rm = TRUE), 2),
    median_gdp = round(median(gdp_nominal.x, na.rm = TRUE), 2),
    min_gdp = round(min(gdp_nominal.x, na.rm = TRUE), 2),
    max_gdp = round(max(gdp_nominal.x, na.rm = TRUE), 2),
    sd_gdp = round(sd(gdp_nominal.x, na.rm = TRUE), 2),
    n = n()
  )
```

Democratic countries tend to have higher GDP levels than non-democratic countries. The mean GDP for democracies (362) is much larger than that of non-democracies (106), and the median is also higher (23.3 vs. 11.4). The large differences in mean and median GDPs between the two groups suggest that democracies tend to perform better economically. However, both groups show substantial variability, with large standard deviations and very high maximum values, indicating the presence of some extreme cases.

```{r}
democracy_data %>%
  filter(!is.na(is_democracy)) %>%
  mutate(is_democracy = factor(is_democracy, labels = c("No", "Yes"))) %>%
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  geom_boxplot(fill = "lightblue", color = "blue", outlier.color = "blue") +
  stat_summary(
    fun = mean,
    geom = "point",
    color = "blue",
    size = 1.5) +
  stat_summary(
    fun = mean,
    geom = "text",
    aes(label = paste0("Mean: ", round(after_stat(y), 1),"%")),
    vjust = -1,
    color = "black") +
  stat_summary(
    fun = median,
    geom = "point",
    color = "blue",
    size = 1.5) +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = paste0("Median: ", round(after_stat(y), 1),"%")),
    vjust = 1.5,
    color = "black") +
  labs(
    title = "Women’s Representation in National Parliament by Democracy Status",
    x = "Democracy Status",
    y = "Percentage of National Parliament Seats Held by Women") + 
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_minimal()
```

Democracies tend to have slightly higher representation of women in national parliaments compared to non-democracies. The median percentage is higher for democracies (15.8%) than for non-democracies (14%), and the mean is also somewhat higher (17.7% vs. 16%).

However, the distributions are fairly similar overall, with substantial overlap between the two groups’ boxplots. Both democracies and non-democracies show wide variability and several outliers. Non-democratic countries seem to have more outliers, including some at higher values, with two outliers exceeding 60%.
